\documentclass[11pt]{article}

% ── Encoding & fonts ──────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}

% ── Page geometry ─────────────────────────────────────────────────────────────
\usepackage[margin=1.25in]{geometry}

% ── Mathematics ───────────────────────────────────────────────────────────────
\usepackage{amsmath}
\usepackage{amssymb}

% ── Tables ────────────────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}

% ── Lists ─────────────────────────────────────────────────────────────────────
\usepackage{enumitem}
\setlist{itemsep=2pt, topsep=4pt}

% ── Code / typewriter text ────────────────────────────────────────────────────
\usepackage{listings}
\lstset{basicstyle=\ttfamily\small, breaklines=true}

% ── Hyperlinks ────────────────────────────────────────────────────────────────
\usepackage[colorlinks=true, linkcolor=black, citecolor=black,
            urlcolor=black, pdfborder={0 0 0}]{hyperref}
\usepackage{url}

% ── Bibliography ──────────────────────────────────────────────────────────────
\usepackage[numbers, sort&compress]{natbib}
\bibliographystyle{unsrtnat}

% ── Paragraph spacing ─────────────────────────────────────────────────────────
\usepackage{parskip}
\setlength{\parskip}{6pt}

% ── Section formatting ────────────────────────────────────────────────────────
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.6em}{}
\titlespacing{\section}{0pt}{14pt}{6pt}
\titlespacing{\subsection}{0pt}{10pt}{4pt}

% ── Abstract ──────────────────────────────────────────────────────────────────
\usepackage{abstract}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small}

% ── Footnotes ─────────────────────────────────────────────────────────────────
\usepackage[hang, flushmargin]{footmisc}

% ─────────────────────────────────────────────────────────────────────────────
\title{\textbf{The Coordinator Trap: Why Mediated Multi-Agent\\
Architectures Cannot Scale and a Substrate-Based Alternative}}

\author{%
  Dr.\ Richard Nicholson\\[4pt]
  \small Tathata Systems Ltd, United Kingdom\\
  \small \href{mailto:tathatasystems@proton.me}{tathatasystems@proton.me}
}

\date{June 2026\\[4pt]
\small Preprint.}
% ─────────────────────────────────────────────────────────────────────────────

\begin{document}

\maketitle

\begin{abstract}
Multi-agent coordination systems are converging on two dominant architectural
patterns: \emph{mediated hierarchies}, in which a central engine aggregates agent
outputs, decides, and issues commands; and \emph{registry-based discovery}, in
which a global index brokers introductions between agents. Both patterns introduce
a coordinator---by different names---that becomes a bottleneck, a single point of
failure, and an unbounded audit obligation as agent populations grow. We
demonstrate by construction that these failure modes are not implementation deficiencies
but structural consequences of the coordinator assumption itself. Drawing on
Holland's framework for Complex Adaptive Systems, we propose an alternative in
which coordination emerges from substrate properties rather than explicit
protocols. We present Mycelium, a working implementation of this model as an
embeddable Rust library, and show how its three-layer architecture---gossip KV
store, signal mesh with boundary admission, and epidemic consensus---structurally
eliminates each failure mode rather than ameliorating it.
\end{abstract}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}

Every major AI agent framework released between 2023 and 2026---LangChain,
AutoGen, CrewAI, OpenAI Swarm, and their successors---converges on the same
structural pattern: a central orchestrator receives requests, routes them to
specialist sub-agents, aggregates their outputs, and returns a synthesised
result~\cite{silobench,mas-survey,raps}. The orchestrator is the coordinator.
The sub-agents are its workers.

This paper argues that the coordinator is not a solution to the multi-agent
coordination problem. It is a restatement of the problem at a different scale.

The coordinator trap produces three failure modes that are structural, not
incidental: (i)~unbounded audit obligation, as every agent output passes through
the coordinator and generates an artifact of equivalent apparent authority
regardless of relevance; (ii)~context loss, as shared state held in or near the
coordinator is lost on restart, forcing agents to re-litigate settled decisions;
and (iii)~output format mismatch, as the coordinator broadcasts undifferentiated
results to heterogeneous consumers it has no model of.

These failure modes have been documented in production deployments. In the
Cisco Mycelium project---an unrelated multi-agent orchestration system---the
coordinator could not maintain context across restarts, produced three-hour
and five-second problems with the same apparent authority, and generated
structured machine logs that failed to communicate to human
reviewers~\cite{signal-noise,context-problem}.

We show that these failure modes are irreducible by improving the coordinator.
They follow necessarily from its existence. We then derive a coordinator-free
architecture from John Holland's signal/boundary model~\cite{holland} and present
Mycelium, a working implementation as an embeddable Rust library.

% ─────────────────────────────────────────────────────────────────────────────
\section{Background: The Two Dominant Patterns}

\subsection{Mediated Hierarchy}

A mediated hierarchy places a cognitive engine at the centre of the agent
population. Specialist agents receive structured payloads from the engine, process
them, and return results. The engine aggregates, decides, and broadcasts. Agent
populations in contemporary LLM frameworks~\cite{mas-survey} follow this pattern:
the orchestrator holds the conversation state and the routing logic; sub-agents
are workers in a fanout RPC system.

Kubernetes~\cite{kubernetes} represents a more mature instance of the same
architectural pattern. A control plane holds the declared target state;
reconciliation loops continuously compare actual cluster state to that target and
issue corrective actions. The coordinator is more sophisticated than a negotiation
engine, but the structural relationship is identical: state is held above the
runtime, and correctness is enforced top-down.

\subsection{Registry-Based Discovery}

A registry-based system separates coordination from discovery. Agents register
capabilities in a global index; other agents query the index to locate peers. The
registry brokers introductions; agents communicate point-to-point thereafter.
MIT~NANDA~\cite{nanda} and FIPA's Directory Facilitator~\cite{fipa,jade} exemplify
this pattern. The registry is a coordinator with a narrower mandate: it holds
authoritative state about agent capabilities and through which discovery passes.

\subsection{The Shared Assumption}

Superficially, these two patterns are different. A mediated hierarchy coordinates
behaviour at runtime; a registry brokers introductions at discovery time. But they
share a single structural assumption: \textbf{there exists a designated component
that holds authoritative state about the agent population and through which
coordination passes.}

This assumption is the coordinator trap. Section~\ref{sec:trap} traces each
failure mode to this assumption directly.

\subsection{The Separation Tax}

The coordinator assumption has a concrete operational cost. A representative
conventional ML inference pipeline must address at minimum four distinct
infrastructure concerns: task dispatch and queuing (Redis/Celery or Kafka
depending on scale), worker registration and discovery (Consul, etcd, or a
sidecar), failure recovery and dead-letter handling, and cross-component
observability. Each concern requires a separately deployed, operated, and
monitored service. The pipeline is not the model---it is the scaffolding around
the model.

Each concern is a separate operational surface: a separate failure mode, a
separate scaling policy, a separate upgrade cycle, and a conceptual translation
layer. The coordination work---route an item to a capable worker, throttle when
workers are saturated, clean up abandoned items, make topology visible---is the
same abstract work at every level. The \emph{separation tax} is the cost of
performing that work redundantly at each concern boundary rather than once in a
unified substrate.

The two dominant patterns described in \S\S2.1--2.2 do not reduce this tax. They
add another coordination system on top of the existing stack. This paper's claim
is that the separation tax is not a necessary cost of multi-agent coordination. It
is the cost of applying the coordinator assumption independently at each layer of
the stack.

% ─────────────────────────────────────────────────────────────────────────────
\section{A History of the Coordinator Assumption}

The coordinator assumption has been the default in distributed computing for over
fifty years. Each generation has recognised the problems it causes and attempted
to mitigate them without eliminating the assumption itself.

\subsection{Blackboard Systems (1970s--80s)}

The blackboard model, most prominently realised in HEARSAY-II~\cite{hearsay},
organises coordination around a central shared data structure and a set of
specialised knowledge sources that read from and write to it. A control component
monitors the blackboard and decides which knowledge source to activate next.
HEARSAY demonstrated that the model could solve hard problems; it also
demonstrated the failure modes: the control component became a bottleneck and a
single point of failure, and the blackboard accumulated stale entries.

\subsection{The Actor Model (Hewitt, 1973)}

Hewitt, Bishop and Steiger's Actor model~\cite{hewitt} provided the first formal
model of concurrent computation as communicating entities. The model eliminates
shared mutable state---a genuine advance---but retains explicit addressing: to
send a message, an actor must know the recipient's address. Topology must be
managed explicitly by the sender. The Actor model correctly identified agents as
the unit of composition; it did not identify boundaries as the mechanism for
self-organising communication.

Mature actor runtimes---most notably Erlang/OTP---address fault tolerance and
process supervision without a central coordinator, and represent a genuine step
beyond the original formulation. The addressing problem persists, however:
Erlang's process registry is a coordinator with a narrower mandate, and OTP
supervision trees impose a coordinator hierarchy for failure propagation. The
addressing problem is resolved not by a more sophisticated registry but by
boundary admission: the emitter declares scope; receivers self-select.

\subsection{Linda and the Tuple Space (Gelernter, 1985)}

Gelernter's Linda~\cite{linda} introduced generative communication via a shared
tuple space. Carriero and Gelernter~\cite{linda-coord} established that
coordination is a distinct language design concern, separable from
computation---a correct and underappreciated insight. What Linda retained,
however, is a shared coordinator: the tuple space itself must remain accessible,
consistent, and available to all participants. Fault-tolerant variants required
replication schemes and consistency protocols. The medium is the coordinator. The
coordinator trap persists in a subtler form.

Carvalho~\cite{carvalho} observes that modern AI agent frameworks are
independently rediscovering Linda's tuple space pattern---implementing ad-hoc
shared stores with task claiming, polling, and matching---without awareness of
forty years of research on its failure modes. The community is not learning from
history. It is repeating it.

\subsection{BDI Agents and FIPA Standards (1990s)}

The Belief-Desire-Intention agent architecture~\cite{bdi} and the Foundation for
Intelligent Physical Agents~\cite{fipa,jade} brought deliberative reasoning to
multi-agent systems. FIPA defined a Directory Facilitator---a registry where
agents register capabilities and query for peers---that is precisely the
registry-based pattern that MIT~NANDA echoes today, four decades later.

\subsection{The Pattern That Persists}

Across five decades, each model correctly identified coordination as a concern
distinct from computation. None eliminated the coordinator assumption:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Model} & \textbf{Coordinator retained} \\
\midrule
Blackboard / HEARSAY (1980) & Central blackboard + control component \\
Actor model (Hewitt, 1973) & Explicit addressing---sender must know recipient \\
Linda (Gelernter, 1985) & Shared tuple space \\
BDI / FIPA / JADE (1990s) & Directory Facilitator as registry \\
Modern LLM orchestration (2020s) & Central orchestrator / CognitiveEngine \\
\bottomrule
\end{tabular}
\end{center}

% ─────────────────────────────────────────────────────────────────────────────
\section{The Coordinator Trap: From Architecture to Symptom}
\label{sec:trap}

We use the term \emph{coordinator trap} to describe the set of failure modes that
any architecture inheriting the coordinator assumption must exhibit. These are not
bugs; they are structural properties. A better coordinator does not escape them.

\subsection{The Mediated Hierarchy Described Precisely}

In a mediated hierarchy, agents are workers in a fanout RPC system. The
coordinator aggregates responses, applies its decision logic, and broadcasts
results. When we call these components ``agents'' we are importing a concept from
the Complex Adaptive Systems literature. Holland~\cite{holland} defines an agent
as an entity that holds a \emph{boundary}---a receptor set specifying the
conditions under which it acts on incoming signals. The boundary is intrinsic to
the agent, not delegated from outside. However, in a mediated hierarchy, agents satisfy
none of this definition: they receive what the coordinator routes, not what they
declare themselves competent to receive. They are passive processors of
coordinator-routed payloads.

\subsection{Audit Burden: The Consequence of Post-Admission Broadcasting}

The unbounded audit obligation follows directly from the coordinator's broadcast
model. Every output of every agent passes through the coordinator and produces a
coordinator-level artifact with identical structure and apparent authority,
regardless of the quality or validity of the underlying agent reasoning. This is
post-admission filtering: artifacts are produced first; quality is assessed after.
As agent populations grow, the coordinator's cognitive load grows linearly with
the number of domains it must evaluate. The audit burden scales with the
coordinator. The three-hour problem and the five-second problem have the same
voice because the coordinator processes them identically.

\subsection{Context Loss: The Consequence of Coordinator-Held State}

In a mediated hierarchy, shared state lives in or near the coordinator. Agents are
stateless workers; they hold no durable view of prior decisions. When the
coordinator restarts, the state is gone. When a new agent joins, it must query the
coordinator for a catchup briefing. The attempted fix documented
in~\cite{context-problem}---a filesystem-based shared memory---correctly identifies
that state should be held outside the coordinator, but introduces a shared mutable
store with no eviction policy: entries persist indefinitely, accumulate, and drift
from current system state. This is not a solution to the coordinator trap; it is a
redistribution of the same problem.

\subsection{Output Format Mismatch: The Consequence of Absent Boundary Model}

When the coordinator synthesises a result and broadcasts it, it has no model of
its recipients' boundaries---no representation of which consumers can act on which
signals. All downstream consumers receive the same output because, from the
coordinator's perspective, they are all equivalent consumers of its decisions. A
human reviewer is not equivalent to an agent as a consumer of a coordination
decision, but the coordinator cannot know this without a model of recipient
boundaries.

\subsection{The Category Error: Workers Dressed as Agents}

The three failure modes share a common root: in a mediated hierarchy, components
called ``agents'' are not agents in any theoretically meaningful sense. A component
whose domain is determined by the coordinator's routing decisions is a worker node
in a fanout RPC system. Calling these components ``agents'' inherits none of the
architectural properties that make genuine agents scalable.

The exit from the coordinator trap requires not a better coordinator but the
elimination of the coordinator from the substrate.

% ─────────────────────────────────────────────────────────────────────────────
\section{Theoretical Foundation: Holland's Signal/Boundary Model}

John Holland's framework for Complex Adaptive Systems~\cite{holland,holland-intro},
formalised in \emph{Signals and Boundaries: Building Blocks for Complex Adaptive
Systems} (MIT Press, 2012), provides the theoretical basis for a coordinator-free
architecture.

Holland's thesis: the behaviour of complex adaptive systems emerges from two and
only two primitives.

\textbf{Signals} propagate through a medium unconditionally. No signal is withheld
from propagation on the basis of who might act on it.

\textbf{Boundaries} are receptor sets. Each agent holds a boundary---a set of
conditions under which it \emph{acts} on a signal. Forwarding is always
unconditional; acting is always conditional on boundary admission.

This inversion is the key insight. Conventional distributed systems thinking
routes messages to known recipients; the emitter must know who is listening. In
Holland's model, any agent whose boundary matches a signal responds. The emitter
does not need to know who is listening. The system tolerates churn without stalling
because there is no routing table to maintain, no coordinator to query.

\textbf{Stigmergy and TTL evaporation.} Holland identifies stigmergy---state left
in the medium by agents, readable by other agents---as the mechanism through which
agents coordinate without direct communication~\cite{dorigo}. Applied to
distributed state management, this suggests that shared state should carry a
time-to-live and evaporate unless actively refreshed. An agent that is alive and
relevant keeps its state fresh; a dead agent's state simply fades. Failure is
emergent: the absence of refreshment \emph{is} the failure signal.

Applied to the failure modes of Section~\ref{sec:trap}: pre-admission filtering
structurally prevents the audit burden---an agent whose boundary does not match a
signal produces no artifact to audit. Distributed TTL state makes
coordinator-type context loss structurally highly improbable---there is no
coordinator memory to lose; state evaporation under simultaneous node failure
remains possible but requires active failure of TTL renewal across all replicas
simultaneously. Self-differentiating boundaries make output format mismatch a
design problem rather than a routing obligation. Each failure mode is not
ameliorated; it is either structurally prevented or made structurally highly
improbable at the substrate level.

% ─────────────────────────────────────────────────────────────────────────────
\section{Prior Art: Correct Concepts, Wrong Implementation}

Four prior systems identified the correct underlying concepts but implemented them
with too much protocol ceremony, or in the wrong deployment model, to achieve
their full potential.

\subsection{Jini and the Lease Insight (Sun Microsystems, 1998)}

Sun Microsystems' Java/Jini~\cite{jini-arch,jini-spec} introduced the insight
that distributed resource registrations should \emph{decay} rather than persist
indefinitely. A service holds a lease on its registration; if it does not renew,
the registration expires. This provides implicit failure detection without
requiring an explicit deregistration protocol. The insight is correct. The
implementation was protocol-heavy: explicit \texttt{Lease} objects,
\texttt{renew()} RPCs, a lease manager, explicit cancellation. The ceremony
obscures the substrate property---that registrations should evaporate unless
actively maintained---behind an explicit lifecycle protocol.

\subsection{OSGi Requirements and Capabilities}

The OSGi Alliance~\cite{osgi} formalised a dependency model in which software
modules declare capabilities they provide and requirements they need; a resolver
matches providers to consumers. The primitive is correct: declarative matching
between providers and consumers, with the resolver handling wiring. What mainstream
OSGi adoption got wrong was treating resolution as static---performed once at
bundle-install time. The resolver ran at deploy time; a module that disappeared at
runtime left a gap with no mechanism for repair. Paremus Service Fabric
addressed this directly.

\subsection{Paremus Service Fabric and the Reconciliation Engine}

Paremus Service Fabric (circa 2010--2015)~\cite{paremus} demonstrated that the
OSGi Requirements and Capabilities model could be applied as a \emph{continuous
runtime} resolver---re-resolving dependencies as services appeared, disappeared,
and changed, adapting the running system accordingly rather than requiring a
redeploy.

To be precise about what Paremus achieved: the R\&C graph was a \emph{declared
target state}. The runtime was continuously monitored against that target, and
deltas were driven back into convergence whenever they appeared. This is closer to
the Kubernetes control loop than to a shared knowledge graph.

It is worth noting Paremus' own architectural trajectory. The platform began as a
Jini runtime with OSGi components, using Jini's service discovery as the
underlying discovery mechanism. Over time, Jini service discovery was replaced
with the OSGi Remote Services specification backed by an in-house gossip-based
implementation---decentralised discovery, TTL-managed registrations, no central
lookup service. Paremus was, in other words, moving in the right direction at the
substrate level. What did not change was the hermetic relationship between the
gossip layer and the reconciliation engine: `as above, so below'---the
declared target state lived in the central engine above; the gossip layer revealed
the landscape below; the engine held the authority to reconcile the two. The
coordinator trap persisted not in the discovery mechanism but in the architectural
layer that held the target.

What Paremus still required was a \emph{central reconciliation engine} holding
that target state. If it went down, reconciliation stopped. The coordinator trap in
a more sophisticated form.

\textbf{The key inversion from Paremus.} In Paremus, the target state was held
\emph{above} the runtime in a central graph and pushed down into components. In
Mycelium, the target state is \emph{compiled into the runtime components
themselves}---each node carries its own fragment as capability and requirement
declarations. There is nothing external to converge toward. Convergence emerges
upward from those fragments.

\subsection{Flow-Based Programming (Morrison, c.~1971)}

Flow-Based Programming~\cite{fbp}, developed by J.~Paul Morrison at IBM in the
early 1970s, conceived programs as networks of independent black-box processes
communicating through bounded connections. The model separates coordination from
computation and gets three things right: (i)~\emph{reusable processes} that know
nothing about their neighbours; (ii)~\emph{automatic backpressure}, in which
bounded connections block writers when full; and (iii)~\emph{declarative topology}
expressed as a data structure separate from component code.

What FBP retained was an external medium to which all these properties belong. The
FBP runtime \emph{is} the coordinator: it allocates connections, manages
backpressure, and resolves the wiring. The coordinator trap remains.

The distinction that Section~\ref{sec:mycelium} realises: in FBP, backpressure is
a runtime-managed property of the connections. An alternative is for each agent to
carry its own opacity declaration as a local KV entry---``I am saturated; treat me
as unavailable''---readable by any resolver in the mesh without a shared runtime
maintaining any state.

\subsection{The Strip-the-Ceremony Pattern}

A pattern emerges across all four cases:

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{>{\bfseries}lXXX}
\toprule
Prior art & Correct concept & Implementation ceremony & Substrate property equivalent \\
\midrule
Jini & Registrations should decay & \texttt{Lease.renew()} RPC + lease manager & TTL as natural evaporation \\
OSGi & Declarative capability matching & Static bundle-install resolver & Continuous evaluation against live state \\
Paremus & Continuous reconciliation toward target state & Central reconciliation engine + target state graph & Gossip mesh --- every TTL refresh is a reconciliation tick \\
FBP & Backpressure, declarative topology, reusable processes & External FBP runtime managing connections & Opacity KV flag --- agent self-declares saturation; resolvers skip it \\
\bottomrule
\end{tabularx}
\end{center}

The pattern that produces better architecture: identify the correct concept, find
the substrate \emph{property} that produces the same behaviour without an explicit
protocol, implement the property and let the behaviour emerge.

% ─────────────────────────────────────────────────────────────────────────────
\section{Mycelium: A Coordinator-Free Substrate}
\label{sec:mycelium}

Mycelium is an embeddable Rust library implementing Holland's signal/boundary
model. It has no daemon, no control plane, no installer, no orchestrator. It is a
crate embedded in the caller's process. The architecture is three layers; each directly addresses one of the three
structural failure modes identified in Section~\ref{sec:trap}: audit burden
(\S4.2), context loss (\S4.3), and output format mismatch (\S4.4).

\subsection{Layer~I --- KV Store: Eliminating Context Loss}

Layer~I is a gossip-replicated key-value store. Every entry carries a Hybrid
Logical Clock timestamp for causal last-write-wins ordering and a time-to-live.
Entries are replicated across the cluster via epidemic gossip; anti-entropy
reconciliation runs on peer reconnection. There is no primary, no leader, no
coordinator. Every node holds a full replica of the key namespace relevant to it.

\textbf{Failure mode eliminated.} Context loss in the mediated hierarchy arises
because shared state is held in or near the coordinator. In Mycelium's Layer~I,
there is no coordinator memory to lose. A node that restarts reconnects to its
peers, runs anti-entropy, and recovers full mesh state within one gossip cycle.

\textbf{TTL evaporation as stigmergy.} Every capability advertisement, every group
membership record, every opacity flag is a KV entry with a TTL. A live node
continuously re-advertises; the trail stays fresh. A dead node stops
re-advertising; its entries evaporate. There is no explicit failure detection
protocol because absence of refreshment \emph{is} the failure signal.

\subsection{Layer~II --- Signal Mesh: Eliminating Audit Burden and Output Mismatch}

Layer~II is an ephemeral signal mesh implementing Holland's signal/boundary model
directly. Signals are emitted with a scope and propagated unconditionally to all
reachable nodes. At each receiving node, \texttt{Boundary::admits()} evaluates the
signal against the node's declared receptor set. Only nodes whose boundary matches
the signal act on it. Forwarding is always unconditional; acting is always
conditional on boundary admission.

\textbf{Audit burden eliminated.} A signal scoped to the \texttt{gpu-compute} group
only activates nodes in that group. An agent in an unrelated group receives and
forwards the signal but produces no artifact. The three-hour problem and the
five-second problem never land on a reviewer who cannot evaluate them because the
boundary prevents admission in the first place.

\textbf{Output format mismatch eliminated.} A signal emitted to a
\texttt{human-review} group reaches only nodes whose boundary includes
human-review membership; a signal emitted to \texttt{automated-pipeline} reaches a
different set. Human-facing and agent-facing consumers self-differentiate via their
boundary declarations without any routing logic in the emitter.

\subsection{Capability and Requirement Declarations: Eliminating the Category Error}

Each Mycelium node holds a \texttt{CapabilityGroupDef}---a declarative
specification of capabilities it provides and requirements it needs. The node
independently evaluates whether it should join each capability group by testing its
own capabilities against the group's filter, right now, against live KV state.
Membership is self-assigned, not coordinator-assigned. The node \emph{is} its
boundary. The application's coordination topology emerges from the composition of
these declarations across the live node population, not from a coordinator's
routing table.

\subsection{Layer~III --- Consensus: Eliminating the Coordinator Bottleneck}

Layer~III provides epidemic consensus---\texttt{group\_propose},
\texttt{system\_propose}, and \texttt{cross\_group\_propose}---implemented entirely
through the signal mesh. \texttt{PROPOSE}, \texttt{VOTE}, and \texttt{COMMIT} are
signal payloads riding ordinary Layer~II frames. The substrate is unaware that
consensus is happening.

\texttt{cross\_group\_propose} supports ballots where multiple named capability
groups act as independent voting blocs. A proposal commits only when every group
independently reaches its required quorum fraction. This supports multi-AZ
durability requirements, compliance ratification, and hierarchical AI pipelines---all
without a coordinator.

\textbf{Failure mode eliminated.} In a mediated hierarchy, coordination throughput
is bounded by the coordinator's capacity, and coordinator failure stops
coordination entirely. In Layer~III, there is no coordinator. No single node's
failure prevents other ballots from proceeding. The system degrades gracefully to
reduced quorum, not coordination failure.

\subsection{Agentic Flow Networks}

The three-layer architecture described in \S\S7.1--7.4 supports a distinct
multi-agent computational pattern that we term an \emph{Agentic Flow Network}
(AFN). An AFN is a topology of capability-bounded agent groups in which work flows
through the mesh as a sequence of KV state transitions, each stage consuming from
one key prefix and producing to the next.

An AFN has five structural properties:

\begin{enumerate}
\item \textbf{Substrate Unity.} All coordination---capability advertisement, group
  membership, backpressure, work items, completion signals---is carried by the same
  gossip KV substrate. There is no separate message queue, registry, or coordination bus.

\item \textbf{Topology Emergence.} Which nodes participate in which pipeline stages
  is not configured by an operator. Each node evaluates its own capabilities against
  the group definitions and self-joins the matching stages.

\item \textbf{Fluid Allocation.} There is no static assignment of workers to
  stages. Any node capable of serving a stage joins its group; the group's size
  fluctuates with available capacity.

\item \textbf{Opacity Backpressure.} A saturated node writes an opacity flag to
  \texttt{sys/load/\{self\}/capacity} (TTL~$\approx$5\,s), making itself invisible
  to resolvers at the next stage. The topology self-throttles without any
  flow-control protocol.

\item \textbf{TTL-Native Cleanup.} Work items written to the KV substrate carry
  TTLs. Abandoned items evaporate automatically. There is no dead-letter queue, no
  explicit cancellation protocol.
\end{enumerate}

These five properties are validated empirically by \texttt{examples/fluid\_pipeline/}
---a 10-worker, 4-stage news article pipeline (fetch $\to$ parse $\to$ score $\to$
publish) running over a shared KV ring as distributed buffer, exercised end-to-end
by integration scenario~11.

An AFN is not a new programming model layered on top of Mycelium. It is the name
for the pattern that naturally arises when a distributed application is built
correctly on a substrate that eliminates the coordinator.

\subsection{The Economic Case: Quadratic Cost Decomposition}

LLM agent execution cost scales superlinearly with iteration count. Because each
API call re-processes the entire accumulated context, the token count at
iteration~$k$ is $\Delta T + \tau k$ (where $\Delta T$ is the initial prompt and
$\tau$ is context growth per iteration). Summing over $k$ iterations yields a
total proportional to $k^2$: cost is quadratic in iteration count, not linear.

The decomposition leverage is direct. Splitting a $k$-iteration task into $M$
independent subtasks of $k/M$ iterations each reduces total cost by factor~$M$:
\[
  M \times \left(\frac{k}{M}\right)^2 = \frac{k^2}{M}
\]
versus $k^2$ for the monolithic run. A 4-way decomposition of a 100-iteration task
costs $4 \times 25^2 = 2{,}500$ versus $100^2 = 10{,}000$---a 4$\times$ saving
from a 4$\times$ decomposition.

A mediated hierarchy cannot realise this saving without explicit orchestration
overhead. The coordinator's synthesis step, where it aggregates $M$ subtask
results into a combined output, recreates the full accumulated context and pays
the quadratic penalty once more. Mycelium's capability-bounded groups implement
decomposition automatically: each group's TTL constrains how long a work item
remains live in that stage, which under typical workloads approximates a bound on
effective $k$ for that group---a time bound rather than a strict iteration count
bound, but one that produces the same practical effect when per-iteration latency
is roughly constant. The quadratic cost accumulation resets at each group
boundary, and no coordinator synthesis is required.

The practical significance of this saving is reinforced by empirical energy
measurements. White~\cite{white2025} demonstrates that inference energy per
query varies by two orders of magnitude across model families and sizes, and
that energy cost is highly sensitive to input characteristics. Where the
quadratic accumulation compounds an already expensive per-call baseline, the
saving from decomposition is not a theoretical nicety---it is an operational
necessity.

% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation}

Mycelium's correctness across its three-layer architecture is validated by 243
unit tests and 12 integration scenarios run against a live 5-node Docker cluster.
Scenarios cover KV replication under partition and reconnection, signal delivery
and boundary admission, capability group formation and dissolution, consensus
quorum under node failure, cross-group voting, the full Agentic Flow Networks
pipeline, and Prompt Skills cross-node KV propagation with LLM invocation. All
12 scenarios pass at HEAD.

The implementation is publicly available at
\url{https://github.com/RichardEko/mycelium} under the AGPL-3.0 licence. The
AGPL-3.0 licence is intentional: applications that deploy Mycelium as part of a
networked service must publish their source under the same terms, closing the
SaaS loophole that permissive licences leave open. Organisations requiring
proprietary embedding may obtain a commercial licence from Tathata Systems Ltd.
The integration test harness is included in the repository and reproducible with a
single \texttt{make test} invocation against a Docker-composed 5-node cluster.

% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{The Hayek Parallel}

The coordinator trap is not a new discovery. Friedrich Hayek described it in
1945---for economies.

In \emph{The Use of Knowledge in Society}~\cite{hayek}, Hayek argued that the
central planning problem is not computational---it is epistemic. No central
planner can possess the distributed, local, tacit knowledge held by individual
market participants. Prices are not just numbers; they are signals that aggregate
and propagate dispersed knowledge through the economy without anyone needing to
understand the whole.

The parallel to the coordinator trap is exact. A planned economy and a mediated
hierarchy share the same failure mode: a central node that must aggregate
knowledge it cannot fully possess, synthesise decisions on behalf of participants
who understand their local context better than any coordinator can, and issue
commands downward into a system whose ground truth is always more current at the
edges than at the centre.

Hayek's market is a signal/boundary system. He just did not have Holland's
vocabulary.

\subsection{The Beinhocker Parallel: Organisations as Complex Adaptive Systems}

Beinhocker's \emph{The Origin of Wealth}~\cite{beinhocker} reaches the same
conclusion from a third direction: organisations. Drawing on extensive empirical
analysis of corporate performance, Beinhocker finds that hierarchical organisations
centralising strategic information and decision-making systematically underperform
distributed ones---not because their leaders are less capable, but because the
local knowledge required for adaptive decisions cannot travel up the chain fast
enough to remain actionable by the time it arrives.

Hayek, Holland, and Beinhocker are describing the same property of complex
adaptive systems at three different scales---economies, biological systems, and
organisations. The coordinator trap is not a software engineering peculiarity. It
is a structural consequence of centralising coordination in any sufficiently
complex adaptive system.

\subsection{Limitations}

Mycelium assumes a cluster the operator owns. Mycelium's scope is intra-cluster;
cross-organisational discovery requires external registry infrastructure.
Mycelium's conforming A2A endpoint makes it reachable from any A2A-speaking
registry without modification to either side.

Ephemeral signals are intentionally not durable---a node that misses a signal
misses it. Durable delivery is a higher-order concern built on the KV layer or
consensus, not a substrate property.

Boundary admission requires agents to declare their boundaries correctly. A
misconfigured boundary---too broad or too narrow---produces incorrect routing
without any coordinator to catch the error. This places a correctness obligation
on the capability declarations that the mediated hierarchy places on the
coordinator instead.

The quadratic cost decomposition in \S7.6 assumes that the $k$-iteration task
can be partitioned into $M$ independent subtasks. Many real agentic workloads
are inherently sequential---multi-step reasoning chains, document drafting,
iterative refinement---where each step depends causally on the prior and
decomposition is not valid. The saving applies where subtask independence holds;
it does not apply universally.

\textbf{When not to use AFN.} Five cases warrant a different approach:
\emph{Strict ordering}---where tasks must be processed in a guaranteed sequence,
not by the first available worker.
\emph{Exactly-once delivery}---where each work item must be processed exactly
once, with no duplicate execution under any failure scenario.
\emph{Complex DAG dependencies}---where later stages must block on combinations
of earlier results in ways that cannot be expressed as linear KV-prefix pipelines.
\emph{Long-term log retention}---where TTL evaporation is a correctness property;
applications requiring replay capability must use a separate append-only store.
\emph{Cross-cluster fan-out at scale}---where the gossip mesh operates within a
single cluster; multi-cluster deployments require a federation layer.

\subsection{Future Work}

\begin{itemize}
\item \textbf{Empirical comparison} against a deployed mediated hierarchy at
  equivalent agent counts. Key measurements: (i)~coordination convergence
  time---single-ballot \texttt{group\_propose} vs NegMAS~\cite{negmas} SAO
  N-round negotiation; (ii)~failure tolerance---coordinator failure in a mediated
  hierarchy vs random node failure in Mycelium; (iii)~state freshness under
  churn---TTL evaporation latency vs knowledge-graph drift rate; (iv)~audit
  obligation under load---artifact count growth as agent population scales from 3
  to 30.

\item \textbf{Formal verification} of the signal/boundary substrate properties
  using TLA+ or similar.

\item \textbf{Cross-cluster federation}---practical experience deploying multiple
  independent Mycelium clusters registered with an internet-scale A2A
  gateway~\cite{nanda}, measuring discovery latency and trust propagation across
  organisational boundaries.

\item \textbf{Signal reorder buffer}---receiver-side per-(sender, kind)
  HLC-keyed causal delivery for applications requiring strict signal ordering.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}

The coordinator is not a solution to the multi-agent coordination problem. It is a
restatement of the problem at a different scale. Mediated hierarchies make the
coordinator responsible for filtering, memory, routing differentiation, and fault
tolerance---properties that in a correctly designed substrate belong to the agents
themselves.

Unbounded audit obligation, context loss across restarts, and output format
mismatch between heterogeneous consumers are not bugs in specific implementations.
They are structural consequences of the coordinator assumption, predictable from
first principles, and irreducible by improving the coordinator.

Holland's signal/boundary model provides the theoretical basis for a different
architecture: one in which coordination emerges from substrate properties rather
than explicit protocols. Mycelium implements this model as an embeddable library.
Each of its three layers---gossip KV store, signal mesh with boundary admission,
epidemic consensus---addresses a mirrored failure mode, not by handling the
failure gracefully but by making each failure mode structurally prevented or,
where substrate constraints apply, structurally highly improbable.

The target state for a Mycelium application is not a document held in a registry
or a graph maintained by a reconciliation engine. It is the aggregate of what
every node declares itself to be, compiled into the runtime components at build
time, assembled bottom-up by the mesh at runtime, and always current because
anything not actively maintained evaporates. Coordination emerges. No coordinator
required.

The coordinator trap is not a new discovery. Hayek described it for economies in
1945. Holland formalised it for complex systems in 2012. Mycelium implements the
solution for AI agent fleets. The insight is the same in all three cases:
distributed local knowledge, expressed through signals and boundaries, produces
emergent order that no coordinator can match.

% ─────────────────────────────────────────────────────────────────────────────
\bibliography{references}

% ─────────────────────────────────────────────────────────────────────────────
\section*{About the Author}

\textbf{Dr.\ Richard Nicholson} is CEO and founder of Tathata Systems Ltd and
Chief AI Transformation Officer at Novus-i2, a strategic transformation company
based in the United Kingdom. He was the founder and CEO of Paremus Ltd., the
company behind Paremus Service Fabric---the continuous OSGi Requirements \&
Capabilities runtime resolver discussed in Section~6.3 as direct conceptual
predecessor to Mycelium's capability model. First-hand experience designing and
deploying Service Fabric over a decade of production use informed both the
architectural critique presented in this paper and the design decisions made to
move beyond it.

He served as President, Board Director, and Treasurer of the OSGi Alliance for
approximately six years~\cite{osgi-mod,osgi-cmb}. He was awarded the OSGi Laureate
(2019) for outstanding contributions to the Alliance and the OSGi Leadership Award
(2019) for service as a board officer and advocate for OSGi modularity strategy.
He was co-architectural lead for the EU Horizon~2020 Brain-IoT project, a
federated platform for emergent coordination across heterogeneous IoT and edge
devices~\cite{brain-iot-1,brain-iot-2}.

\end{document}
